Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-31 survey/survey.tex
revised sec 3 a bit more up to wavefront

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data I: Finite Sets}

@@

Instead of the spaces consisting solely of unstructured isolated
points, we now investigate structured spaces built from sums and
products of elementary spaces. This structure corresponds to the
building blocks of type theory which are: the empty type
($\mathbf{0}$), the unit type ($\mathbf{1}$), the sum type ($\uplus$),
and the product ($\times$) type. Before getting into the formal
theory, let's consider possible deformations on the space
$(\mathbf{1} \uplus \mathbf{0}) \times (\mathbf{1} \uplus
\mathbf{1})$. This space is the product of two subspaces: the subspace
$(\mathbf{1} \uplus \mathbf{0})$ which itself is the sum of the space
$\mathbf{1}$ containing one element and the empty space $\mathbf{0}$
and the subspace $(\mathbf{1} \uplus \mathbf{1})$ which is the sum of
two spaces each containing one element. First, as discussed in the
previous section, any deformation of this space must at least preserve
the number of elements: we can neither create nor destroy points
during any continuous deformation. Seeing that the number of elements
in our example space is 2, a reasonable hypothesis is that we can
deform the space above to any other space with 2 elements. What this
really means is that we are treating the sum and product structure as
malleable. For example, imagining a product structure as arranged in a
grid, by stretching we can turn this structure to a sum structure
arranged in a line.

We now turn into the mathematical formalization of this idea. Our goal
is a sound and complete equivalence relation on types that maintains
the number of points. First we note that the type structure of types
has a nice correspondence (Curry-Howard) to logic:

\begin{center}
\begin{tabular}{c|c}
Logic & Types \\ \hline
$\false$ & $\mathbf{0}$ \\
$\true$ & $\mathbf{1}$ \\
$\land$ & $\times$ \\
$\lor$ & $\uplus$ \\
\end{tabular}
\end{center}

This correspondence is rather fruitful. As logical expressions form a
commutative semiring, we would expect that types too form a
commutative semiring. And indeed they do -- at least up to \emph{type
  isomorphism}.  The natural numbers $\Nat$ are another commutative
semiring; it will turn out that, even though the Curry-Howard
correspondence has been extremely fruitful for programming language
research, it is $\Nat$ which will be a better model for finite
structured types as the corresponding commutative semiring captures
the familiar numerical identities that evidently preserve the number
of points in the types.

\begin{defn}
  A \emph{commutative semiring} (sometimes called a \emph{commutative
    rig} (commutative ri\emph{n}g without negative elements) consists of a
  set $R$, two distinguished elements of $R$ named 0 and 1, and two
  binary operations~$+$ and $\cdot$, satisfying the following
  relations for any $a,b,c \in R$:
\begin{equation}
\begin{array}{rcl}
0 + a &=& a \\
a + b &=& b + a \\
a + (b + c) &=& (a + b) + c \\
\\
1 \cdot a &=& a \\
a \cdot b &=& b \cdot a \\
a \cdot (b \cdot c) &=& (a \cdot b) \cdot c \\
\\
0 \cdot a &=& 0 \\
(a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
\end{array}
\label{eq:csemiring}
\end{equation}
\end{defn}

@@

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set''.  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\mathbf{0} : U$ (and similarly for the other
items). Our first instinct would be to similarly replace $=$ with a
judgement $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal}. This is however not true: unless
$A \equiv B$, $A \times B \equiv B \times A$ is not normally
provable\footnote{Not without assuming that $U$ is univalent, which we
  don't.}. But it should be clear that $A \times B$ and $B \times A$
contain equivalent information. In other words, we would like to be
able to witness that $A \times B$ can be reversibly deformed into
$B \times A$. Which motivates the introduction of type
\emph{equivalences}. To do this, we need a few important auxiliary
concepts first.

\begin{defn}[homotopy]
\label{def:homotopy}
Two functions $f,g:A \rightarrow B$ are \emph{homotopic} if
$\forall x:A. f(x) = g(x)$. We denote this $f \sim g$.
\end{defn}

\noindent It is easy to prove that homotopies (for any given function
space $A \rightarrow B$) are an equivalence relation.  The simplest
definition of the data which makes up an equivalence is then

@@

\noindent Why \emph{quasi}? The reasons are beyond our scope, but
the interested reader can read sections $2.4$ and $4$ in \cite{hottbook}.
There are several conceptually different, but equivalent,
``better'' definitions.  We record just one here:

@@

If we replace $=$ by $\simeq$, then we can indeed prove that
types (with $\mathbf{0}, \mathbf{1}, \uplus, \times$) form a commutative semiring.

The reader familiar with universal algebra should pause
and ponder a bit about what we have done. We have lifted
\emph{equality} from being in the signature of the ambient logic
and instead put it in the signature of the algebraic structure
of interest. Then, and only then, are we in a solid position
to show that we have a concept (commutative semiring) which is flexible
enough for types to be an instance.

If we revisit the Curry-Howard correspondence, we notice one
more issue. In logic, it is true that $A \lor A = A$ and
$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
equivalent to $A$. They are however \emph{equi-inhabited}. This is
a fancy way of saying
\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow
  \qquad A \ \text{is inhabited} \]
The above is the real \textit{essence} of the correspondence.
In other words, classical Curry-Howard tells us about
\emph{logical equivalence} of types. This is even a
constructive statement: there are indeed functions
$f : A \uplus A \rightarrow A$ and $g : A \rightarrow A \uplus A$;
however, they are not inverses.

So mere inhabitation falls far short of our goals of being
able to smoothly deform from one type to another.

Let us thus analyze the crux of the ``problem''. In logic, we have
that $\land$ and $\lor$ are both \emph{idempotent}: this is the
property of any binary operation $*$ where
$\forall a. a * a = a$. And it should be clear that idempotent
operations are \emph{forgetful}: its input has two copies of $a$,
but its output, only one. On the type side, something more subtle
happens. Consider $\mathbf{1} \uplus \mathbf{1}$ versus $\mathbf{1}$; the first has
\emph{two} proofs of inhabitation ($\texttt{inl tt} and \texttt{inr tt}$)
while the second only one ($\texttt{tt}$). These cannot be put in
bijective correspondence. Even though the ``payload'' \texttt{tt}
is the same, forgetting $\texttt{inl}$ (or \texttt{inr}) throws
away information -- something we have expressly disallowed.
Yes, this should remind you of Maxwell's demon: even though the
data is the same, they are tagged differently, and this
data is indeed information, and must be preserved.

\begin{figure}[t]
\[
\begin{array}{rrcll}
\idc :& \tau & \simeq & \tau &: \idc \\
\\
\identlp :&  \mathbf{0} \uplus \tau & \simeq & \tau &: \identrp \\
\swapp :&  \tau_1 \uplus \tau_2 & \simeq & \tau_2 \uplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \uplus (\tau_2 \uplus \tau_3) & \simeq & (\tau_1 \uplus \tau_2) \uplus \tau_3 &: \assocrp \\
\\
\identlt :&  \mathbf{1} \times \tau & \simeq & \tau &: \identrt \\
\swapt :&  \tau_1 \times \tau_2 & \simeq & \tau_2 \times \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \times (\tau_2 \times \tau_3) & \simeq & (\tau_1 \times \tau_2) \times \tau_3 &: \assocrt \\
\\
\distz :&~ \mathbf{0} \times \tau & \simeq & \mathbf{0} ~ &: \factorzl \\
\dist :&~ (\tau_1 \uplus \tau_2) \times \tau_3 & \simeq & (\tau_1 \times \tau_3) \uplus (\tau_2 \times \tau_3)~ &: \factor
\end{array}
\]
\caption{Type isomorphisms.}
\label{type-isos}
\end{figure}

Nevertheless, the Curry-Howard correspondence still has some
force. We know that the inhabitants of types formed with
with $\mathbf{0}, \mathbf{1}, \uplus, \times$ are a commutative semiring. What we
want to know is, which types are equivalent? From a commutative semiring
perspective, this amounts to asking what terms are equal?
We have a set of generators for those equations, namely those
in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
of mutually inverse functions which witness these identities.
For concreteness, we show the signatures in Figure~\ref{type-isos}.
As these all come in symmetric pairs (some of which are self-symmetric),
we give names for both directions. The names use $\iso$, $+$ and $*$
as these will be what we use when we reflect these isomorphism into
a programming language, which we proceed to do next.

@@

We now have in our hands our desired denotational semantics: we want
to create a language $\Pi$ such that the types and type combinators
map to $\mathbf{0}, \mathbf{1}, \uplus, \times$, and such that we have ground
terms whose denotation are all $16$ type isomorphisms of
Figure~\ref{type-isos}. This is completely straightforward, as
we can simply do this literally. To make the analogy with
commutative semirings stand out even more, we will use $0, 1, +$, and $*$
at the type level, and will denote ``equivalence'' by $\iso$.
Thus Figure~\ref{pi-terms} shows the ``constants'' of the language.

@@

Of course, one does not get a programming language with just typed
constants! There is a need to perform multiple equivalences. There are
in fact three ways to do this: sequential composition $\odot$, 
choice composition $\oplus$ (sometimes called juxtaposition) and parallel
composition $otimes$. See Figure~\ref{pi-combinators} for the signatures.

$c_1 \odot c_2$ corresponds to performing $c_1$ first, then $c_2$, and
is the usual notion of composition -- and corresponds to $\fatsemi$ of
the language of the previous section. $c_1 \oplus c\_2$ chooses to
perform $c_1$ or $c_2$ depending on whether the input is labelled
$\textsf{inl}$ or $\textsf{inr}$ respectively. $c_1 \otimes c_2$
operates on a product structure, and applies $c_1$ to the first component
and $c_2$ to the second.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements} We would like to thank the numerous
students and colleagues who participated in various aspects of this
research and who provided valuable feedack and constructive criticism.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data I: Finite Sets}
\label{sec:dataone}

@@

Instead of spaces consisting solely of unstructured isolated points,
we now investigate structured spaces built from sums and products of
elementary spaces. This structure corresponds to the building blocks
of type theory which are: the empty type ($\bot$), the unit type
($\top$), the sum type ($\uplus$), and the product ($\times$)
type. Before getting into the formal theory, let's consider possible
deformations on the space
$(\top \uplus \bot) \times (\top \uplus \top)$. This space is the
product of two subspaces: the subspace $(\top \uplus \bot)$ which
itself is the sum of the space $\top$ containing one element
$\texttt{tt}$ and the empty space $\bot$ and the subspace
$(\top \uplus \top)$ which is the sum of two spaces each containing
the one element $\texttt{tt}$. First, as discussed in the previous
section, any deformation of this space must at least preserve the
number of elements: we can neither create nor destroy points during
any continuous deformation. Seeing that the number of elements in our
example space is 2, a reasonable hypothesis is that we can deform the
space above to any other space with 2 elements such as
$\top \uplus \top$ or $\top \uplus (\top \uplus \bot)$. What this
really means is that we are treating the sum and product structure as
malleable. For example, imagining a product structure as arranged in a
grid, by stretching we can turn this structure to a sum structure
arranged in a line, change the orientation of the grid by exchanging
the axes, as well as do other transformations to be discussed in this
section that preserve the number of points.

%%%%%%%%%
\subsection{A Model of Type Equivalences}

We now turn into the mathematical formalization of this idea. Our goal
is a denotational semantics on types that maintains the number of
points. First we note that the type structure of types has a nice
correspondence (Curry-Howard) to logic:

\begin{center}
\begin{tabular}{c|c}
Logic & Types \\ \hline
$\false$ & $\bot$ \\
$\true$ & $\top$ \\
$\land$ & $\times$ \\
$\lor$ & $\uplus$ \\
\end{tabular}
\end{center}

\noindent This correspondence is rather fruitful. As logical
expressions form a commutative semiring, we would expect that types
too form a commutative semiring. And indeed they do -- at least up to
\emph{type isomorphism}.  The natural numbers $\Nat$ are another
commutative semiring; it will turn out that, even though the
Curry-Howard correspondence has been extremely fruitful for
programming language research, it is $\Nat$ which will be a better
model for finite structured types as the corresponding commutative
semiring captures the familiar numerical identities that preserve the
number of points in the types.

\begin{defn}
\label{def:rig}
A \emph{commutative semiring} (sometimes called a \emph{commutative
  rig} (commutative ri\emph{n}g without negative elements)
$(R,0,1,+,\cdot)$ consists of a set $R$, two distinguished elements of
$R$ named 0 and 1, and two binary operations~$+$ and $\cdot$,
satisfying the following relations for any $a,b,c \in R$:
\[\begin{array}{rcl}
0 + a &=& a \\
a + b &=& b + a \\
a + (b + c) &=& (a + b) + c \\
\\
1 \cdot a &=& a \\
a \cdot b &=& b \cdot a \\
a \cdot (b \cdot c) &=& (a \cdot b) \cdot c \\
\\
0 \cdot a &=& 0 \\
(a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
\end{array}\]
\end{defn}

@@

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set''.  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\bot : U$ (and similarly for the other
items). Our next instinct would be to similarly replace $=$ with a
judgement $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal}. This is however not true: the
proposition $A \times B \equiv B \times A$ is not
normally\footnote{Except In homotopy type theory where equivalent
  types are identified.} provable for arbitrary types $A$ and $B$. But
it should be clear that $A \times B$ and $B \times A$ contain
equivalent information. In other words, we would like to be able to
witness that $A \times B$ can be reversibly deformed into
$B \times A$. Which motivates the introduction of type
\emph{equivalences}. To do this, we need a few important auxiliary
concepts.

\begin{defn}[Homotopy]
\label{def:homotopy}
Two functions $f,g:A \rightarrow B$ are \emph{homotopic} if
$\forall x:A. f(x) \equiv g(x)$. We denote this $f \sim g$.
\end{defn}

\noindent It is easy to prove that homotopies (for any given function
space $A \rightarrow B$) are an equivalence relation.  The simplest
definition of the data which makes up an equivalence is the following.

@@

\noindent Why \emph{quasi}? The reasons are beyond our scope, but the
interested reader can read Sec.~$2.4$ and Ch.~$4$ in the HoTT
book~\cite{hottbook}.  There are several conceptually different, but
equivalent, ``better'' definitions.  We record just one here:

@@

We are now in position to formally state the appropriate commutative
semiring structure for types. After replacing the set $R$ with a
universe $U$, we also replace the algebraic use of $=$ in
Def.~\ref{def:rig} by the type equivalence relation $\simeq$. With
this change, we can indeed prove that types (with
$\bot, \top, \uplus, \times$) form a commutative semiring. The reader
familiar with universal algebra should pause and ponder a bit about
what we have done. We have lifted \emph{equality} from being in the
signature of the ambient logic and instead put it in the signature of
the algebraic structure of interest. Then, and only then, are we in a
solid position to show that we have a concept (commutative semiring)
which is flexible enough for types to be an instance.

If we revisit the Curry-Howard correspondence, we notice one
more issue. In logic, it is true that $A \lor A = A$ and
$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
equivalent to $A$. They are however \emph{equi-inhabited}. This is
a fancy way of saying
\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow \qquad A \
  \text{is inhabited} \] The above is the real \textit{essence} of the
Curry-Howard correspondence.  In other words, classical Curry-Howard
tells us about \emph{logical equivalence} of types. This is even a
constructive statement: there are indeed functions
$f : A \uplus A \rightarrow A$ and $g : A \rightarrow A \uplus A$;
however, they are not inverses.

So mere inhabitation falls far short of our goals of being able to
smoothly deform from one type to another. Let us thus analyze the crux
of the ``problem''. In logic, we have that $\land$ and $\lor$ are both
\emph{idempotent}: this is the property of any binary operation $*$
where $\forall a. a * a = a$. And it should be clear that an
idempotent operations is a \emph{forgetful} operation: its input has
two copies of $a$, but its output, only one. On the type side,
something more subtle happens. Consider $\top \uplus \top$
versus $\top$; the first has \emph{two} proofs of inhabitation
($\texttt{inl tt}$ and $\texttt{inr tt}$) while the second only one
($\texttt{tt}$). These cannot be put in bijective correspondence. Even
though the ``payload'' \texttt{tt} is the same, forgetting
$\texttt{inl}$ (or \texttt{inr}) throws away information -- something
we have expressly disallowed.  Yes, this should remind you of
Maxwell's demon: even though the data is the same, they are tagged
differently, and this data is indeed information, and must be
preserved.

\begin{figure}[t]
\[
\begin{array}{rrcll}
\idc :& A & \simeq & A &: \idc \\
\\
\identlp :&  \bot \uplus A & \simeq & A &: \identrp \\
\swapp :&  A \uplus B & \simeq & B \uplus A &: \swapp \\
\assoclp :&  A \uplus (B \uplus C) & \simeq & (A \uplus B) \uplus C &: \assocrp \\
\\
\identlt :&  \top \times A & \simeq & A &: \identrt \\
\swapt :&  A \times B & \simeq & B \times A &: \swapt \\
\assoclt :&  A \times (B \times C) & \simeq & (A \times B) \times C &: \assocrt \\
\\
\distz :&~ \bot \times A & \simeq & \bot ~ &: \factorzl \\
\dist :&~ (A \uplus B) \times C & \simeq & (A \times C) \uplus (B \times C)~ &: \factor
\end{array}
\]
\caption{Type isomorphisms.}
\label{type-isos}
\end{figure}

Nevertheless, the Curry-Howard correspondence still has some force. We
know that the inhabitants of types formed with with
$\bot, \top, \uplus, \times$ are a commutative semiring. What we want
to know is, which types are equivalent? From a commutative semiring
perspective, this amounts to asking what terms are equal?  We have a
set of generators for those equations, namely those in
Def.~\ref{def:rig}. What we thus need is to create $8$ pairs of
mutually inverse functions which witness these identities.  For
concreteness, we show the signatures in Fig.~\ref{type-isos}.  As
these all come in symmetric pairs (some of which are self-symmetric),
we give names for both directions.

%%%%%%%%%
\subsection{A Language of Type Equivalences}

@@

We now have in our hands our desired denotational semantics. We want
to create a programming language, which we call $\Pi$, such that the
types and type combinators map to $\bot, \top, \uplus, \times$, and
such that we have ground terms whose denotation are all $16$ type
isomorphisms of Fig.~\ref{type-isos}. This is completely
straightforward, as we can simply do this literally. To make the
analogy with commutative semirings stand out even more, we will use
$0, 1, +$, and $\times$ at the type level, and will denote
``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
``constants'' of the language.

@@

Of course, one does not get a programming language with just typed
constants! There is a need to perform multiple equivalences. There are
in fact three ways to do this: sequential composition $\odot$, choice
composition $\oplus$ (sometimes called juxtaposition), and parallel
composition $\otimes$. See Fig.~\ref{pi-combinators} for the
signatures. The construction $c_1 \odot c_2$ corresponds to performing
$c_1$ first, then $c_2$, and is the usual notion of composition -- and
corresponds to $\fatsemi$ of the language of permutations of
Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c\_2$ chooses to
perform $c_1$ or $c_2$ depending on whether the input is labelled
$\textsf{inl}$ or $\textsf{inr}$ respectively. Finally the
construction $c_1 \otimes c_2$ operates on a product structure, and
applies $c_1$ to the first component and $c_2$ to the second.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements} We would like to thank the numerous
students and colleagues who participated in various aspects of this
research and who provided valuable feedback and constructive criticism.
